---
title: "\\textbf{Modern Salary Modeling Project}"
author: "Justin Mai, Troy Russo, Isaac Muhlestein, Conan Li, Jian Kang"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document: default
geometry: margin=37pt
fontsize: 11pt
header-includes:
  - \usepackage{titling}
  - \usepackage{titlesec}
  - \titlespacing*{\title}{0pt}{0pt}{0pt}
  - \setlength{\droptitle}{-2em}
  - \setlength{\topskip}{0pt} 
mainfont: "Times New Roman"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
```

---

# 1. Introduction

**Description:** The job market can be a hard place to navigate, especially with the search of data roles in the recent years. As statistics students, many of us are leaning towards opportunities within data roles. To understand the recent market we will be analyzing data job logistics to investigate the factors and predictors that most impacts the salary of these roles. Within this report, we will be using Salary Index data reported by real people in the industry to (1) discover the factors and variables within the job description that may influence a person's job salary the most to help students like us navigate the market and (2) ...

**Disclaimer:** We have pivoted from using this dataset https://www.kaggle.com/datasets/uom190346a/ai-powered-job-market-insights which is comprised of synthetic data based off the current job market regarding AI jobs to this dataset https://www.kaggle.com/datasets/murilozangari/jobs-and-salaries-in-data-field-2024/data which consists of real survey data from various people in data roles, reporting through this website https://aijobs.net/salaries/2024/. We decided to make this change because we believe that variables such as `experience_level` and `job_category` which can be found in our current dataset would be strong predictors for `salary`. We also believe that using real survey data as supposed to synthetic data would give us results that are more related to real-life circumstances, making the report more applicable for all.

---

# 2. Methods

## 2.1 Data Description

```{r include=FALSE}
job_data <- read.csv("jobs_in_data_2024.csv")

job_data <- job_data %>% 
  select(!c(work_year, salary_currency, salary))

cost_of_living <- read.csv("cost_of_living_2024.csv")

cost_of_living <- cost_of_living %>% 
  rename(employee_residence = Country)
```

The first dataset was collected through https://aijobs.net/salaries/2024/, it consists of 14199 different observations, with each observation representing a person in their role in 2024. The **response variable** we are measuring is `salary_in_usd` which measures a person's annual gross salary. The **8 predictors** are `experience_level`, `employment_type`, `job_title`, `employee_residence`, `work_setting`, `company_location`, `company_size`, `job_category`. All of these variables are categorical where `company_size` is categorized as *S* for small, *M* for medium, and *L* for large.

The second dataset consists of cost of living index by country where an index of 100 represents the living cost of NYC, United States, so all the indices are relative to that. We will merge the two datasets by `country`. The predictors we're looking at in this dataset are Cost of Living Index, Rent Index, Cost of Living Plus Rent Index, and Local Purchasing Power Index. We believe that the cost of living could be indicative of `salary_usd`.

## 2.2 Data Processing

The primary dataset will be comprised of the two datasets described in (2.1). We are joining the two datasets on `employee_residence` which is in form of country. Now each row will consist of a specified job description along with the cost indexes for each respective resident. Having all of these predictors in one dataset will allow us to utilize the lm() function to uncover linear trends for all predictor variables in response to `salary`. It will also allow us to compare models easily which we will do using ANOVA tests and by calculating the F-statistic. The primary dataset consists of 14199 observations after joining

**Data Manipulation**: Rows that consisted of NAs were in countries that weren't listed in the `cost_of_living` data, this demonstrates that their rank is low when ordering by index and there weren't a sufficient number of samples for those countries. Therefore we removed those observations (14161 observations). We also removed exact duplicate rows from the dataset (7575 observations)

Mutations in the data were also made to create new predictors `us_resident` which is a binary variable that denotes if the job is in the U.S. or not, and `experience_numeric` which turns `experience_level` into numerical values (i.e. 1 - "Entry-Level", 2 - Mid-level", 3 - "Senior", 4 - "Executive"), this transformation will support our use of linear modeling and allow us to easily check assumptions such as linearity assumptions. Also, because we have too many different job titles, we decided to aggregate these job titles by keywords into 8 categories (Data Scientist, Data Analyst, Machine Learning, Data Engineer, Leadership, Business Intelligence, Research, Other). This will consolidate our data and make linear models more interpretable

We are also reordering values to ensure that our **baseline term** is what we want it to be (i.e. releveling small companies to be the first type and entry-level jobs to be the first job types). 

```{r echo=FALSE}
# changing categorical to as.factor
job_data <- job_data %>%
  mutate(across(where(is.character), as.factor))

# Joining the two datasets
joined_df <- job_data %>% 
  left_join(cost_of_living, by=c("employee_residence"))

# removing NAs
joined_df <- na.omit(joined_df)

# new variable
joined_df <- joined_df %>% 
  mutate(us_resident = ifelse(employee_residence == "United States", 1, 0))

# changing experience level to a numeric
joined_df <- joined_df %>%
  mutate(experience_numeric = case_when(
    experience_level == "Entry-level" ~ 1,
    experience_level == "Mid-level" ~ 2,
    experience_level == "Senior" ~ 3,
    experience_level == "Executive" ~ 4
  ))

# ordering levels
joined_df$company_size <- factor(joined_df$company_size, levels = c("S", "M", "L"))
joined_df$experience_level <- factor(joined_df$experience_level, levels = c("Entry-level", "Mid-level", "Senior", "Executive"))

# aggregating job titles to job categories
joined_df <- joined_df %>%
  mutate(job_category = case_when(
    grepl("Data Scientist|Data Science|Integration|Applied Scientist", job_title, ignore.case = TRUE) ~ "Data Scientist",
    grepl("Analyst|Analytics|Modeler", job_title, ignore.case = TRUE) ~ "Data Analyst",
    grepl("Machine Learning|ML|AI", job_title, ignore.case = TRUE) ~ "Machine Learning",
    grepl("Engineer|Architect|Developer", job_title, ignore.case = TRUE) ~ "Data Engineer",
    grepl("Manager|Director|Lead|Head|Management", job_title, ignore.case = TRUE) ~ "Leadership",
    grepl("Business Intelligence|BI", job_title, ignore.case = TRUE) ~ "Business Intelligence",
    grepl("Research", job_title, ignore.case = TRUE) ~ "Research",
    TRUE ~ "Other"
  ))

# removing duplicate rows
joined_df <- joined_df %>% 
  distinct()
```

## 2.3 Model Diagnostics

**Linear Modeling Assumptions**:

\begin{itemize}
    \item \textbf{Linearity}: The relationship between the predictor and response is linear.
    \item \textbf{Independence}: All observations are independent of one another (pair-wise independence).
    \item \textbf{Homoscedasticity}: The variance of residuals is constant across predictor levels.
    \item \textbf{Residual Normality}: The residuals follow a normal distribution.
\end{itemize}

We know that the reported job descriptions are all independent of one another through the data description.

The model that we are using seems highly categorical, even after our data transformation process, which would result in very discrete predictions. To fix this issue we want to create a new interaction term and turn it into a predictor variable, adding a continuous predictor for `salary_usd`. We hypothesize that salary growth will vary by location, therefore, we are creating an interaction term between `experience_numeric` and the Cost of Living Index to test this (experience combined with living costs could impact salary growth differently).

The residuals when using a non-transformed model is skewed due to the deviations within the tails in our qq-plot, to fix this issue, we would have to use a **log-transformation** on the data. The variance of the residuals is also constant and there is a clear linear and positive relationship between the predictor and response.

```{r message=FALSE, echo=FALSE, fig.width=5, fig.height=3, fig.align='center', fig.cap='Normality check for Experience and Cost of Living Index Interaction variable on Salary, comparison between no transformation and log-transformation'}
# mutating data to create interaction term
joined_df <- joined_df %>% 
  mutate(exp_cost_int = experience_numeric * Cost.of.Living.Index)

par(mfrow = c(1, 2))
plot(lm(salary_in_usd ~ exp_cost_int, data = joined_df), which = 2)
title("No Transformation")
plot(lm(log(salary_in_usd) ~ log(exp_cost_int), data = joined_df), which = 2)
title("Log-Transformation")
par(mfrow = c(1, 1))  
```

To test the assumptions for our categorical variables we must look at the average salary by category, to see if there is a clear trend between the different experience levels in respective company size. This checks off the linearity assumption because there is a (somewhat) linear increasing trend for all the experience levels and we can see that adding the company size variable and the necessity of an interaction effect between `experience_level` and `company_size` because although the trend is increasing for each company size, they are increasing at a separate rate

```{r message=FALSE, echo=FALSE, fig.width=5, fig.height=3, fig.align='center', fig.cap='Interaction between Experience Level and Company Size on Salary'}
mean_salary <- joined_df %>%
  group_by(experience_level, company_size) %>%
  summarize(mean_salary = mean(salary_in_usd))

ggplot(mean_salary, aes(x = experience_level, y = mean_salary, color = company_size, group = company_size)) +
  geom_point(size = 3) +
  geom_line() +
  labs(title = "Mean Salary by Experience Level and Company Size") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 3. Results

## 3.1 Data Exploration

```{r echo=FALSE, warning=FALSE}
joined_df %>% 
  ggplot(aes(x = experience_level, y = salary_in_usd)) + 
  geom_boxplot() + 
  facet_wrap(~ us_resident, labeller = labeller(us_resident = c("0" = "Non-U.S. Resident", "1" = "U.S. Resident"))) + 
  labs(title = "Salary Distribution by Experience Level",
       x = "Experience Level",
       y = "Salary (USD)")
```

## 3.2 Model Selection

To determine our model, we will fit multiple linear regression models and find the model that minimizes our **AIC (Akaike Information Criterion) and/or BIC (Bayesian Information Criterion)**. If the AIC and BIC suggest different models, we will favor the model selected by lowest AIC because BIC penalizes models with a large number of observations and tends to predict less than the AIC.

We can use a step function from the MASS package that computes the best model purely based on the AIC by comparing every potential model combination and returning the model with the lowest AIC. This model may be overfit however, since AIC does not account for model complexity.
```{r, echo = F, eval = F}
# Load the MASS package for stepAIC
library(MASS)

# Fit the full model with all predictors
full_model <- lm(salary_in_usd ~ ., data = joined_df)

# Perform stepwise selection using AIC (both directions)
step_model <- stepAIC(full_model, direction = "both", trace = FALSE)

# Display the summary of the selected model
summary(step_model)


```
Now that we have the model that we have found from AIC we can test the assumptions of the model as well as performing cross validation to test the valididity of the model. We can also perform an anova test to see if our predictors are significant.
```{r}
# Assuming your final model is stored in 'step_model' and your data frame is 'joined_df'

# Alternatively, you can plot individual diagnostic plots:
# Residuals vs Fitted
plot(step_model$fitted.values, step_model$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

# Q-Q Plot for Normality of Residuals
qqnorm(step_model$residuals, main = "Q-Q Plot")
qqline(step_model$residuals, col = "red")

# Testing for heteroscedasticity using the Breusch-Pagan test
library(lmtest)
bp_test <- bptest(step_model)
print(bp_test)  # p-value < 0.05 indicates potential heteroscedasticity

# 2. Model Validation: Cross-Validation -----------------------------

# Using the caret package for 10-fold cross-validation
# Install caret if needed: install.packages("caret")
library(caret)
set.seed(123)  # for reproducibility

# Define training control for 10-fold cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Refit the model using caret's train() function. 
# Here, we use the same predictors that were selected in your step_model.
cv_model <- train(salary_in_usd ~ experience_level + job_title + employee_residence + 
                  work_setting + company_size, 
                  data = joined_df, 
                  method = "lm", 
                  trControl = train_control)
print(cv_model)


# 3. Theoretical Considerations -------------------------------------
# Although not strictly "code," here are some steps to integrate theory:

# - Review each predictor’s significance and consider if domain knowledge suggests its retention.
#   For example, even if certain levels of a categorical predictor have high p-values, the variable might still be
#   important overall. You could run an ANOVA to test the overall significance of categorical variables:
anova_result <- anova(step_model)
print(anova_result)
```
The cross validation tells us that our model is not over fitting since the $R^2$ is only about 35%, but also suggests that we can improve our model because the RMSE is quite high at 54993, so we may need to add more interactions or do some non linear transformations of the data.

The residual plot shows a funnel shape, which suggests hetereoscadascity of the variance and the residuals do not appear to be centered at zero, which indicates that there is bias. The plots indicate that we may want to use the log of the salary instead of salary as our response.

we can now run the step AIC function again using the log(salary) as our response and perform the same model diagnostic tests as above. We also add an interaction term between experience_level and company size.

```{r}
joined_df$log_salary <- log(joined_df$salary_in_usd)
log_full_model <- lm(log_salary ~ ., data = joined_df)
step_log_model <- stepAIC(log_full_model, direction = "both", trace = FALSE)
summary(step_log_model)

model_with_interaction <- update(step_log_model, 
                                 . ~ . + experience_level:company_size)

# 2. Perform stepwise selection again starting from the updated model
step_log_model_interact <- stepAIC(model_with_interaction, 
                                   direction = "both", 
                                   trace = FALSE)

# 3. Review the summary of the new model
summary(step_log_model_interact)


plot(step_log_model_interact$fitted.values, step_log_model_interact$residuals,
     main = "Residuals vs Fitted",
     xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, col = "red")

qqnorm(step_log_model_interact$residuals, main = "Q-Q Plot")
qqline(step_log_model_interact$residuals, col = "red")

bp_test <- bptest(step_log_model_interact)
print(bp_test)  # p-value < 0.05 indicates potential heteroscedasticity

set.seed(123)  # for reproducibility
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(log_salary ~ experience_level + job_title + employee_residence + 
                  work_setting + company_size, 
                  data = joined_df, 
                  method = "lm", 
                  trControl = train_control)
print(cv_model)
anova_result <- anova(step_log_model_interact)
print(anova_result)
```
Our log transformed model has a much higher $R^2$ indicating that our model explains a lot more of the variability in log(salary) than it does salary. Although our residual plot still shows bias and heteroscedasity, so we need to consider what other transformations we can make.

```{r}
library(MASS)
bc <- boxcox(lm(log_salary ~ .+ experience_level:company_size, data = joined_df), plotit = TRUE)
# Choose lambda based on the plot and refit using transformed response:
lambda_opt <- bc$x[which.max(bc$y)]
joined_df$trans_salary <- if(lambda_opt == 0) log(joined_df$salary_in_usd) else (joined_df$salary_in_usd^lambda_opt - 1)/lambda_opt

```
The box cox indicates that the log transformed model is the best power model for our data, so the changes we need to make will be on indiviual predictors and not the entire data.


## 3.3 Modeling

## 3.5 Model Descriptions

---

# 4. Conclusion