---
title: "\\textbf{Modern Salary Modeling Project}"
author: "Justin Mai, Troy Russo, Isaac Muhlestein, Conan Li, Jian Kang"
output:
  pdf_document:
    toc: true
    latex_engine: xelatex
  html_document: default
geometry: margin=37pt
fontsize: 11pt
header-includes:
  - \usepackage{titling}
  - \usepackage{titlesec}
  - \titlespacing*{\title}{0pt}{0pt}{0pt}
  - \setlength{\droptitle}{-2em}
  - \setlength{\topskip}{0pt} 
mainfont: "Times New Roman"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(tidyverse)
```


---

# Introduction

**Description:** Our goal with this Modern Salary Modeling Project is to analyze modern-day jobs (most are related to AI) to attempt to make salary predictions, looking at the influence of different variables on job salaries. This notebook will use the F-statistic and a series of ANOVA tests (hypothesis and confidence intervals)to determine the best predictors for the job market of the future.

---

# Methods

## Data Description

## Data Manipulation

## Research Questions

## Model Selection

---

# Results

## Data Exploration

## Modeling

### Model Diagnostics

### Model Descriptions

---

# Conclusion